---
title: "GroupProject2"
author: "Tommy Papesh, Jasper Drumm, Aidan Hatzer, Jonah Cuenca, and Leif Gullstad"
date: "2022-10-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Explanation of Data and Business Case
The Data set we chose to use is called “NFL.CSV,” which details the players and player
characteristics of everyone who entered the 2009 through 2019 NFL Drafts. Our business
question that we hope to answer with our model is: Should a Junior in college declare early for
the NFL draft or stay for their senior year? We believe there are many questions that we can
build off of this main question such as figuring out which round the player will likely be selected
in and what their first contract may look like. Our model hopes to consider specific predictor
variables, such as age, school, height, weight, position, and combine performance to predict if a
player is likely to get drafted (our response variable) and help them decide on their future.



## Data Reading and Cleaning
```{r}
nfl <- read.csv("NFL.csv", stringsAsFactors = TRUE)
# get rid of player name since that is basically like the id column we usually get rid of
# also get rid of the Drafted Information since that will impact our model and is a post-condition of the response variable
nfl$Drafted..tm.rnd.yr. <- NULL
nfl$Player <- NULL
nfl$School <- NULL # removing because there is too many low frequency unique values to run logistic regression

# Set Response variable to 0 and 1
nfl$Drafted <- ifelse(nfl$Drafted == "Yes", 1, 0)

# Changing NA variables to the mean
nfl$Age <- ifelse(is.na(nfl$Age), mean(nfl$Age, na.rm = T), nfl$Age)
nfl$Sprint_40yd <- ifelse(is.na(nfl$Sprint_40yd), mean(nfl$Sprint_40yd, na.rm = T), nfl$Sprint_40yd)
nfl$Bench_Press_Reps <- ifelse(is.na(nfl$Bench_Press_Reps), mean(nfl$Bench_Press_Reps, na.rm = T), nfl$Bench_Press_Reps)
nfl$Vertical_Jump <- ifelse(is.na(nfl$Vertical_Jump), mean(nfl$Vertical_Jump, na.rm = T), nfl$Vertical_Jump)
nfl$Broad_Jump <- ifelse(is.na(nfl$Broad_Jump), mean(nfl$Broad_Jump, na.rm = T), nfl$Broad_Jump)
nfl$Agility_3cone <- ifelse(is.na(nfl$Agility_3cone), mean(nfl$Agility_3cone, na.rm = T), nfl$Agility_3cone)
nfl$Shuttle <- ifelse(is.na(nfl$Shuttle), mean(nfl$Shuttle, na.rm = T), nfl$Shuttle)
summary(nfl)
```


## Logistics Model Function
```{r}
LM_func <- function(cluster_norm) {
  library(class)
  library(caret)
  library(gmodels)
  set.seed(12345)
  cluster_sample <- sample(1:nrow(cluster_norm), floor(0.3 * nrow(cluster_norm))) # 30% size of cluster_norm to use for sample
  cluster_test <- cluster_norm[-cluster_sample,] # 70% for test
  cluster_train <- cluster_norm[cluster_sample,]
  
  cluster_model <- glm(Drafted ~ . , data = cluster_train, family = "binomial")
  
  glm_Prediction <- predict(cluster_model, newdata = cluster_test, type = "response")
  glm_Prediction <- ifelse(glm_Prediction < 0.5, 0, 1)
  summary(glm_Prediction)

  CrossTable(x = cluster_test$Drafted, y = glm_Prediction, prop.chisq = F)
  confusionMatrix(as.factor(cluster_test$Drafted), as.factor(glm_Prediction))
}
```

## ANN function
```{r}

ANN_func <- function(cluster_norm) {
  # cluster_norm$cluster <- NULL
  set.seed(12345)
  cluster_sample <- sample(1:nrow(cluster_norm), floor(0.3 * nrow(cluster_norm))) # 30% size of cluster_norm to use for sample
  cluster_test <- cluster_norm[-cluster_sample,] # 70% for test
  cluster_train <- cluster_norm[cluster_sample,]
  
  library(caret)
  library(neuralnet)
  
  func_model <- neuralnet(formula = Drafted~., data = cluster_train, hidden = 2)
  
  func_prediction <- predict(func_model, newdata = cluster_test, type = "response")
  func_prediction <- ifelse(func_prediction < 0.5, 0, 1)
  confusionMatrix(as.factor(cluster_test$Drafted), as.factor(func_prediction))
}
```

## Normalization Function
```{r}
normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}
```


## Normalize numeric rows
```{r}
nfl_norm <- nfl
nfl_norm$Age <- normalize(nfl$Age)
nfl_norm$Height <- normalize(nfl$Height)
nfl_norm$Weight <- normalize(nfl$Weight)
nfl_norm$Sprint_40yd <- normalize(nfl$Sprint_40yd)
nfl_norm$Vertical_Jump <- normalize(nfl$Vertical_Jump)
nfl_norm$Bench_Press_Reps <- normalize(nfl$Bench_Press_Reps)
nfl_norm$Broad_Jump <- normalize(nfl$Broad_Jump)
nfl_norm$Agility_3cone <- normalize(nfl$Agility_3cone)
nfl_norm$Shuttle <- normalize(nfl$Shuttle)
nfl_norm$BMI <- normalize(nfl$BMI)
nfl_norm$Year <- normalize(nfl$Year)
```

## Running Preliminary Logistic Regression
```{r}
# Error being thrown due to the schools, but hopefully we can figure it out together and ask for help if not.
LM_func(nfl_norm)
# ANN_func(nfl_norm)
```

